---
title: "Simulated example"
author: "THM"
date: "August 30, 2019"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(AER)

```

## Simulate correlated data

We simulate data from the model

\begin{align}
y_1 &= 1 -0.1y_2 + \epsilon_1 +\epsilon_2 &(1)\\
y_2 &= w_1 + 0.5\epsilon_1 +\epsilon_3  &(2)\\
z &= w_2 + 1.5\epsilon_3 &(3)
\end{align}

$\epsilon_i$s and $w_i$s are all independent standard normal.

$\epsilon_1$ in (1) and (2) is the source of correlation between $y_2$ and the error term $\epsilon_1+\epsilon_2$ for $y_1$.

$\epsilon_3$ in (2) and (3) provides the correlation between $y_2$ and $z$. 

```{r}

sim.df = tibble(w1 = rnorm(3000), w2 = rnorm(3000), e1 = rnorm(3000), e2 = rnorm(3000), e3 = rnorm(3000))

sim.df = sim.df %>% mutate(z = w2 + 1.5*e3, y2 = w1 + 0.5*e1 + e3, y1 = 1-0.1*y2 + e1 + e2 )

```

## OLS

Let us just do OLS immediately on the data

```{r}
m.OLS = lm(y1~y2, data = sim.df)
summary(m.OLS)
```

Here we see that the intercept is `r m.OLS$coefficients[1]` and the coefficient of $y_2$ is `r m.OLS$coefficients[2]`. Both are also highly significant. 


Let us check the covariance/correlation of $y_2$ with the residuals.
```{r}
sim.df$res = m.OLS$residuals
cov.y2res = cov(sim.df$y2, sim.df$res)
cor.y2res = cor(sim.df$res, sim.df$y2)
cov.y2res
cor.y2res
```

We can see that covariance of $y_2$ with the residuals is `r cov.y2res` and the correlation is `r cor.y2res`. 

## 2SLS

```{r}
m.2SLS = ivreg(y1 ~ y2 | z, data = sim.df)
summary(m.2SLS, diagnostics = TRUE)
```

We see that 2SLS will recover the true estimate of the coefficient of $y_2$, even though the $R^2$ is negative.
