---
title: "Exercise 1"
author: "Tan Hong Ming"
date: "August 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading the data

Here we read the excel file and check if the data read is correct using the head and tail function. Note we used excel preliminary to convert the .xls filetype to .csv. 
```{r}
married.women.df = read.csv('csvExample1.csv')
head(married.women.df)  # check data
tail(married.women.df)  # check data
```

## Plotting the data

For illustrative purposes. Plotting is done using the "ggplot2" package. Documentation can be found at https://ggplot2.tidyverse.org/
Note: You may need to install the package first by using "install.packages("ggplot2")".

```{r}
library(ggplot2)  # load the ggplot2 package
ggplot(married.women.df) + geom_point(aes(ln.earn., ln.hours., colour = AGE, shape=factor(NKIDS)))
```

## Calculating the reduced forms

We use the lm function to perform the least sqaures approximation, then summary function to output a summary of the estimated model.

### Wage

```{r}
wage.reduced.form = lm(ln.earn. ~ NKIDS + AGE, data = married.women.df)
summary(wage.reduced.form)
```

### Hours

```{r}
hours.reduced.form = lm(ln.hours. ~ NKIDS + AGE, data = married.women.df)
summary(hours.reduced.form)
```

## Structural 

```{r}
structural.form = lm(ln.hours. ~ ln.earn., data = married.women.df)
summary(structural.form)
```

## Get wage predictions 

Here we use our least squares approximation model of the wage to get predicted wages. We will use the predicted wages as an instrumental variable in the second step of the two stage least squares. 

```{r}
p.df = predict(wage.reduced.form, married.women.df)
head(p.df)
tail(p.df)
married.women.df$predWage = p.df  # add the predicted values as a new column in our data frame 
```

After this, we perform the second least squares approximation.

```{r}
struc.2sls.form = lm(ln.hours. ~ predWage, data = married.women.df)
summary(struc.2sls.form)
```

## Testing for overindentification restrictions

First, we use our second stage model to predict hours, and calculate the prediction error

```{r}
p.df = predict(struc.2sls.form, married.women.df)
head(p.df)
tail(p.df)
married.women.df$error = married.women.df$ln.hours.- p.df
```

Next, we regress errors from second stage on the instrumental variables.

```{r}
sargan.test.model = lm(error ~ NKIDS + AGE, data = married.women.df)
summary(sargan.test.model)
```

Now we calculate our test-statistic.

```{r}
test.stat = summary(sargan.test.model)$r.squared * nrow(married.women.df) # summary(sargan.test.model)$df[2]
p.value = pchisq(test.stat, 1, lower.tail = FALSE)
```
|test statistic|p value|
|--------------|-------|
|`r test.stat` | `r p.value`|

## Direct two stage least squares with diagnostic tests

We can also perform the two stage least squares directly using R. For this, we use the ivreg() function in the AER package. 
Note: you may need to install the package first by using "install.packages("AER")".

```{r warning=FALSE, message=FALSE}
library(AER)  # load the AER package
direct.2sls.form = ivreg(ln.hours. ~ ln.earn.| NKIDS + AGE, data = married.women.df)
summary(direct.2sls.form, diagnostics = TRUE)
```
